---
title: "Not-So-Great-Expectations. Code for data preprocessing"
author: "Rocío Galeote"
date: "2025-03-31"
output: html_document
---

## Data processing

NOTE: This rmd serves as Part 1 of the accompanying code for the thesis "Not-so-Great Expectations: Analyzing Gender, Genre and Reader Bias in Commercial Fiction Using Natural Language Processing".

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/goodreads.gz"
output_file <- "goodreads.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
goodreadscsv <- "goodreads.csv"

# Read CSV as a tibble
goodreads <- read_csv(goodreadscsv)

# Print summary
summary(goodreadscsv)


```




```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/goodreads2.gz"
output_file <- "goodreads2.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
goodreads2csv <- "goodreads2.csv"

# Read CSV as a tibble
goodreads2 <- read_csv(goodreads2csv)

# Print summary
summary(goodreads2csv)


```




```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/goodreads3.gz"
output_file <- "goodreads3.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
reviews <- "goodreads3.csv"

# Read CSV as a tibble
reviews <- read_csv(goodreads3csv)

# Print summary
summary(reviews)


```

user_id,book_id,review_id,rating,review_text,date_added,date_updated,read_at,started_at,n_votes,n_comments

Data cleaning:

```{r}
library(dplyr)

goodreads_clean <- goodreads3 |> 
  select(-date_added, -date_updated, -read_at, -started_at)
```


Data about books


```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/goodreadsbooks.gz"
output_file <- "goodreadsbooks.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
goodreadsbooks <- "goodreadsbooks.csv"

# Read CSV as a tibble
goodreadsbooks <- read_csv(goodreadsbooks)

# Print summary
summary(goodreadsbooks)


```

Cleaning books:

```{r}
books <- goodreadsbooks |> 
  select(-default_description_language_code, -default_chaptering_book_id, -original_publication_day, -original_publication_month, -original_language_id)

summary(books)
```



```{r}
library(dplyr)

# Count NAs in each column
na_counts <- goodreadsbooks %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  t()  # Transpose for better readability

# Convert to a tibble for a clean output
na_summary <- tibble(Variable = rownames(na_counts), Missing_Values = as.vector(na_counts))

# Print the summary
print(na_summary)
```



```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/goodreadsauthors.gz"
output_file <- "goodreadsauthors.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
authors <- "goodreadsauthors.csv"

# Read CSV as a tibble
authors <- read_csv(authors)

# Print summary
summary(authors)


```

Mystery dataset

```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/mystery.gz"
output_file <- "mystery.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
mystery <- "mystery.csv"

# Read CSV as a tibble
mystery <- read_csv(mystery)

# Print summary
summary(mystery)


```

Cleaning mystery:

```{r}
mystery <- mystery |> 
  select(-date_added, -date_updated, -read_at, -started_at)

summary(mystery)
```



Romance reviews

```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/romancereviews.gz"
output_file <- "romancereviews.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

```

Romance books info:

```{r}

library(jsonlite)
library(dplyr)
library(tidyr)

# Define the file path
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/romancebooks.gz"
output_file <- "romancebooks_cleaned.csv"

# Open the gz file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks of data
json_data <- list()

# Set chunk size (adjustable based on memory capacity)
chunk_size <- 10000  # Process 10,000 lines at a time
counter <- 0

# Process the file in chunks
while (length(lines <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  chunk_data <- lapply(lines, function(x) tryCatch(fromJSON(x), error = function(e) NULL))
  
  # Remove any invalid entries (NULL)
  chunk_data <- chunk_data[!sapply(chunk_data, is.null)]
  
  # Store the parsed data in the json_data list
  json_data <- append(json_data, chunk_data)
  
  # Track progress
  counter <- counter + length(lines)
  cat("Processed", counter, "lines...\n")
}

# Close the connection
close(con)

# Convert the collected data to a dataframe
df <- bind_rows(lapply(json_data, function(x) {
  tibble(
    book_id = x$book_id,
    work_id = x$work_id,
    title = x$title,
    title_without_series = x$title_without_series,
    average_rating = as.numeric(x$average_rating),
    ratings_count = as.numeric(x$ratings_count),
    text_reviews_count = as.numeric(x$text_reviews_count),
    num_pages = as.numeric(x$num_pages),
    publication_year = as.numeric(x$publication_year),
    publication_month = as.numeric(x$publication_month),
    publication_day = as.numeric(x$publication_day),
    language_code = x$language_code,
    isbn = x$isbn,
    isbn13 = x$isbn13,
    asin = x$asin,
    kindle_asin = x$kindle_asin,
    publisher = x$publisher,
    format = x$format,
    description = x$description,
    is_ebook = x$is_ebook,
    link = x$link,
    url = x$url,
    image_url = x$image_url,
    similar_books = paste(x$similar_books, collapse = ", "),  # Convert list to a string
    authors = paste(x$authors$author_id, collapse = ", "),  # Extract author IDs as comma-separated string
    popular_shelves = paste(x$popular_shelves$name, collapse = ", ")  # Extract shelf names
  )
}))

# Show summary of the final dataset
summary(df)

library(tidyverse)
# Save the cleaned dataset as a CSV
write_csv(df, output_file)
cat("Saved cleaned dataset as", output_file, "\n")


```


Mystery reviews:

```{r}

library(jsonlite)
library(data.table)

# Define file paths
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/mysteryreviews.gz"
output_file <- "mysteryreviews.csv"

# Open the compressed file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks
data_list <- list()
chunk_size <- 10000  # Adjust this based on memory (10,000 lines per chunk)

counter <- 0
while (length(line <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  json_objects <- lapply(line, fromJSON)
  
  # Convert to data.table for efficient processing
  df <- rbindlist(json_objects, fill = TRUE)
  
  # Write to CSV in append mode (efficient for large files)
  fwrite(df, output_file, append = TRUE, col.names = (counter == 0))
  
  counter <- counter + 1
  cat("Processed", counter * chunk_size, "lines...\n")
}

# Close the file connection
close(con)

cat("Conversion complete! CSV saved as", output_file)

library(tidyverse)

# Define file path
mystery <- "mystery.csv"

# Read CSV as a tibble
mystery <- read_csv(mystery)

# Print summary
summary(mystery)

```


Mystery books info:

```{r}

library(jsonlite)
library(dplyr)
library(tidyr)

# Define the file path
input_file <- "C:/Users/rochi/Documents/UNIVERSIDAD/MASTER/TFM/mysterybooks.gz"
output_file <- "mysterybooks_cleaned.csv"

# Open the gz file for reading
con <- gzfile(input_file, "rt")

# Initialize an empty list to store chunks of data
json_data <- list()

# Set chunk size (adjustable based on memory capacity)
chunk_size <- 10000  # Process 10,000 lines at a time
counter <- 0

# Process the file in chunks
while (length(lines <- readLines(con, n = chunk_size, warn = FALSE)) > 0) {
  # Parse each line as JSON
  chunk_data <- lapply(lines, function(x) tryCatch(fromJSON(x), error = function(e) NULL))
  
  # Remove any invalid entries (NULL)
  chunk_data <- chunk_data[!sapply(chunk_data, is.null)]
  
  # Store the parsed data in the json_data list
  json_data <- append(json_data, chunk_data)
  
  # Track progress
  counter <- counter + length(lines)
  cat("Processed", counter, "lines...\n")
}

# Close the connection
close(con)

# Convert the collected data to a dataframe
df <- bind_rows(lapply(json_data, function(x) {
  tibble(
    book_id = x$book_id,
    work_id = x$work_id,
    title = x$title,
    title_without_series = x$title_without_series,
    average_rating = as.numeric(x$average_rating),
    ratings_count = as.numeric(x$ratings_count),
    text_reviews_count = as.numeric(x$text_reviews_count),
    num_pages = as.numeric(x$num_pages),
    publication_year = as.numeric(x$publication_year),
    publication_month = as.numeric(x$publication_month),
    publication_day = as.numeric(x$publication_day),
    language_code = x$language_code,
    isbn = x$isbn,
    isbn13 = x$isbn13,
    asin = x$asin,
    kindle_asin = x$kindle_asin,
    publisher = x$publisher,
    format = x$format,
    description = x$description,
    is_ebook = x$is_ebook,
    link = x$link,
    url = x$url,
    image_url = x$image_url,
    similar_books = paste(x$similar_books, collapse = ", "),  # Convert list to a string
    authors = paste(x$authors$author_id, collapse = ", "),  # Extract author IDs as comma-separated string
    popular_shelves = paste(x$popular_shelves$name, collapse = ", ")  # Extract shelf names
  )
}))

# Show summary of the final dataset
summary(df)

library(tidyverse)
# Save the cleaned dataset as a CSV
write_csv(df, output_file)
cat("Saved cleaned dataset as", output_file, "\n")


```

Reading the files


```{r}


library(tidyverse)

# ROMANCE REVIEWS

romancereviews <- "romancereviews.csv"
romancereviews <- read_csv(romancereviews)
summary(romancereviews)

# ROMANCEBOOKS
romancebooks <- "romancebooks_cleaned.csv"
romancebooks <- read_csv(romancebooks)
summary(romancebooks)

#MYSTERY REVIEWS
mysteryreviews <- "mysteryreviews.csv"
mysteryreviews <- read_csv(mysteryreviews)
summary(mysteryreviews)

#MYSTERY BOOKS
mysterybooks <- "mysterybooks_cleaned.csv"
mysterybooks <- read_csv(mysterybooks)
summary(mysterybooks)

```

Cleaning:

```{r}

library(dplyr)

mysteryreviews <- mysteryreviews |> 
  select(-date_added, -date_updated, -read_at, -started_at)

romancereviews <- romancereviews |> 
  select(-date_added, -date_updated, -read_at, -started_at)

mysterybooks <- mysterybooks |> 
  select(-publication_year, -publication_month, -publication_day, -asin, -kindle_asin, -format)

romancebooks <- romancebooks |> 
  select(-publication_year, -publication_month, -publication_day, -asin, -kindle_asin, -format)



# Keep only rows where language_code starts with "en"
mysterybooks_en <- mysterybooks %>%
  filter(grepl("^en(-|$)", language_code))

# Check result
unique(mysterybooks_en$language_code)

# Keep only rows where language_code starts with "en"
romancebooks_en <- romancebooks %>%
  filter(grepl("^en(-|$)", language_code))

# Check result
unique(romancebooks_en$language_code)

write_csv(romancebooks_en, "romancebooks_en.csv")
write_csv(mysterybooks_en, "mysterybooks_en.csv")

```


Saving workspace

```{r}

save(romancebooks, file = "romancebooks.RData")
save(romancereviews, file = "romancereviews.RData")
save(mysterybooks, file = "mysterybooks.RData")
save(mysteryreviews, file = "mysteryreviews.RData")

```


Merging

```{r}

# Left join: keep all reviews, even if the book info is missing
mystery_combined <- left_join(mysteryreviews, mysterybooks_en, by = "book_id")

# Left join: keep all reviews, even if the book info is missing
romance_combined <- left_join(romancereviews, romancebooks_en, by = "book_id")

```

NAs

```{r}

colSums(is.na(mystery_combined))
colSums(is.na(romance_combined))


library(dplyr)

# For mystery_combined
mystery <- mystery_combined |> 
  filter(!is.na(title))

# For romance_combined
romance <- romance_combined |> 
  filter(!is.na(title))

write_csv(romance, "romance.csv")
write_csv(mystery, "mystery.csv")

```

Add authors:

```{r}

authors <- "./2/goodreadsauthors.csv"
authors <- read_csv(authors)
summary(authors)

# How many author IDs in 'mystery' are found in 'authors'?
sum(mystery$authors %in% authors$author_id)

# What proportion of author IDs from 'mystery' exist in 'authors'?
mean(mystery$authors %in% authors$author_id)

mystery$authors <- as.character(mystery$authors)
authors$author_id <- as.character(authors$author_id)
romance$authors <- as.character(romance$authors)

library(dplyr)

mysteryaut <- left_join(mystery, authors, by = c("authors" = "author_id"))
romanceaut <- left_join(romance, authors, by = c("authors" = "author_id"))

```

Final cleaning:

```{r}
library(dplyr)

# Create a cleaned version of mysteryaut
mystery_clean <- mysteryaut %>%
  transmute(
    title = title,
    title_without_series = title_without_series,
    author = name,                    # Rename 'name' to 'autor'
    description = description,
    pages = num_pages,              # Rename 'num_pages' to 'pages'
    language_code = language_code,
    review_id = review_id,
    review_rating = rating,         # Rename 'rating' to 'review_rating'
    review_text = review_text
  )

# Preview the result
glimpse(mystery_clean)


library(dplyr)


romance_clean <- romanceaut %>%
  transmute(
    title = title,
    title_without_series = title_without_series,
    author = name,                    # Rename 'name' to 'autor'
    description = description,
    pages = num_pages,              # Rename 'num_pages' to 'pages'
    language_code = language_code,
    review_id = review_id,
    review_rating = rating,         # Rename 'rating' to 'review_rating'
    review_text = review_text
  )

# Preview the result
glimpse(romance_clean)

write_csv(romance_clean, "romancefin.csv")
write_csv(mystery_clean, "mysteryfin.csv")

```

