---
title: "Not-So-Great-Expectations. Code for H2 testing"
author: "Rocío Galeote"
date: "2025-08-11"
output: html_document
---

## Hypothesis 2 testing:

NOTE: This rmd serves as Part 4 of the accompanying code for the thesis "Not-so-Great Expectations: Analyzing Gender, Genre and Reader Bias in Commercial Fiction Using Natural Language Processing".

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

```{r}
library(dplyr)
library(stringr)
library(tidyr)
library(sentimentr)
library(rcompanion)
```


We create a dataset with only those reviews that include the authors name or the words "author" or "writer".

```{r}

# Helper to safely escape regex special characters
escape_regex <- function(x) {
  str_replace_all(x, "([.|()\\^{}+$*?]|\\[|\\])", "\\\\\\1")
}

# Create a new dataset with author mentions or keywords
mystery_author_mentions <- mystery_final %>%
  mutate(
    author_first = word(author, 1),
    author_last = word(author, -1),
    # Escape names for regex use
    author_first_clean = escape_regex(str_to_lower(author_first)),
    author_last_clean = escape_regex(str_to_lower(author_last))
  ) %>%
  filter(
    str_detect(str_to_lower(review_text), "\\bauthor\\b") |
    str_detect(str_to_lower(review_text), "\\bwriter\\b") |
    str_detect(str_to_lower(review_text), author_first_clean) |
    str_detect(str_to_lower(review_text), author_last_clean)
  )

# View result count
cat("📝 Total reviews mentioning the author or writer:", nrow(mystery_author_mentions), "\n")

# Preview a few rows
head(mystery_author_mentions %>% select(review_id, author, review_text), 5)


# Check gender distribution of the reviews that mention the author or writer
mystery_author_mentions %>%
  count(gender) %>%
  mutate(percentage = round(n / sum(n) * 100, 2)) %>%
  arrange(desc(n))


# Count unique authors per gender
mystery_author_mentions %>%
  distinct(author, gender) %>%
  count(gender)

# Create a new dataset with author mentions or keywords
romance_author_mentions <- romance_final %>%
  mutate(
    author_first = word(author, 1),
    author_last = word(author, -1),
    # Escape names for regex use
    author_first_clean = escape_regex(str_to_lower(author_first)),
    author_last_clean = escape_regex(str_to_lower(author_last))
  ) %>%
  filter(
    str_detect(str_to_lower(review_text), "\\bauthor\\b") |
    str_detect(str_to_lower(review_text), "\\bwriter\\b") |
    str_detect(str_to_lower(review_text), author_first_clean) |
    str_detect(str_to_lower(review_text), author_last_clean)
  )

# View result count
cat("📝 Total reviews mentioning the author or writer:", nrow(mystery_author_mentions), "\n")

# Preview a few rows
head(romance_author_mentions %>% select(review_id, author, review_text), 5)


# Check gender distribution of the reviews that mention the author or writer
romance_author_mentions %>%
  count(gender) %>%
  mutate(percentage = round(n / sum(n) * 100, 2)) %>%
  arrange(desc(n))

# Count unique authors per gender
romance_author_mentions %>%
  distinct(author, gender) %>%
  count(gender)


# Combine gender counts for both datasets, including unknown or NA
author_mentions_summary <- bind_rows(
  mystery_author_mentions %>%
    mutate(gender = if_else(is.na(gender), "unknown", gender)) %>%
    count(dataset = "Mystery", gender, name = "count"),

  romance_author_mentions %>%
    mutate(gender = if_else(is.na(gender), "unknown", gender)) %>%
    count(dataset = "Romance", gender, name = "count")
) %>%
  group_by(dataset) %>%
  mutate(
    total_mentions = sum(count),
    percentage = round(100 * count / total_mentions, 2)
  ) %>%
  ungroup()

# Pivot wider to create a table with counts and percentages
author_mentions_table <- author_mentions_summary %>%
  pivot_wider(
    id_cols = dataset,
    names_from = gender,
    values_from = c(count, percentage),
    values_fill = 0
  ) %>%
  relocate(starts_with("count"), .after = dataset) %>%
  mutate(
    total_mentions = count_female + count_male + count_unknown
  )

# View final table
print(author_mentions_table)

```



Then, we create a function to extract the desired adjectives:

```{r}

extract_adj_dependent_udpipe <- function(parsed, first_name, last_name) {
  if (is.null(parsed) || nrow(parsed) == 0) return(NA_character_)
  
  library(dplyr)
  target_tokens <- tolower(c(first_name, last_name, paste(first_name, last_name), "author", "writer"))
  
  # Find token ids of author references (NOUN or PROPN)
  author_token_ids <- parsed %>%
    filter(tolower(token) %in% target_tokens,
           upos %in% c("PROPN", "NOUN")) %>%
    pull(token_id)
  
  if (length(author_token_ids) == 0) return(NA_character_)
  
  adj_modifiers <- c()
  
  for (idx in author_token_ids) {
    # Find adjectives that *directly* modify the author token (amod or acl relations)
    adjs <- parsed %>%
      filter(head_token_id == idx,
             upos == "ADJ",
             !grepl("NumType=Ord", feats),
             dep_rel %in% c("amod", "acl")) %>%
      pull(token)
    
    # Find negations attached to these adjectives (neg relation)
    neg_adjs <- parsed %>%
      filter(dep_rel == "neg",
             head_token_id %in% (parsed %>% 
                                  filter(head_token_id == idx, upos == "ADJ") %>% 
                                  pull(token_id))) %>%
      pull(head_token_id)
    
    # Add negated adjectives with "not"
    negated_adjs <- parsed %>%
      filter(token_id %in% neg_adjs) %>%
      pull(token)
    if (length(negated_adjs) > 0) {
      adj_modifiers <- c(adj_modifiers, paste("not", negated_adjs))
    }
    
    # Add normal adjectives
    adj_modifiers <- c(adj_modifiers, adjs)
  }
  
  if (length(adj_modifiers) == 0) return(NA_character_)
  return(paste(unique(adj_modifiers), collapse = "; "))
}
  

mystery_author_mentions <- mystery_author_mentions %>%
  rowwise() %>%
  mutate(author_adjectives_ud = extract_adj_dependent_udpipe(udpipe_df, first_name, last_name)) %>%
  ungroup() %>%
  relocate(author_adjectives_ud, .after = udpipe_df)


romance_author_mentions <- romance_author_mentions %>%
  rowwise() %>%
  mutate(author_adjectives_ud = extract_adj_dependent_udpipe(udpipe_df, first_name, last_name)) %>%
  ungroup() %>%
  relocate(author_adjectives_ud, .after = udpipe_df)
```

List to see some of them along with the author gender and name:

```{r}

# Create the adjective list for the mystery genre
mystery_adjective_list <- mystery_author_mentions %>%
  # Select the necessary columns
  select(author, gender, author_adjectives_ud) %>%
  
  # Remove rows where no adjectives were found
  filter(!is.na(author_adjectives_ud)) %>%
  
  # Split the adjective strings into separate rows
  separate_rows(author_adjectives_ud, sep = "; ") %>%
  
  # Rename for clarity
  rename(
    adjective = author_adjectives_ud,
    gender = gender,
    author = author
  ) %>%
  
  # Clean up whitespace
  mutate(adjective = trimws(adjective))

# View the result for the mystery genre
print(mystery_adjective_list)
```

```{r}
# Create the adjective list for the romance genre
romance_adjective_list <- romance_author_mentions %>%
  # Select the necessary columns
  select(author, gender, author_adjectives_ud) %>%
  
  # Remove rows where no adjectives were found
  filter(!is.na(author_adjectives_ud)) %>%
  
  # Split the adjective strings into separate rows
  separate_rows(author_adjectives_ud, sep = "; ") %>%
  
  # Rename for clarity
  rename(
    adjective = author_adjectives_ud,
    gender = gender,
    author = author
  ) %>%
  
  # Clean up whitespace
  mutate(adjective = trimws(adjective))

# View the result for the romance genre
print(romance_adjective_list)

write.csv(romance_adjective_list, "romance_adjective_list.csv", row.names = FALSE)
write.csv(mystery_adjective_list, "mystery_adjective_list.csv", row.names = FALSE)

```


Now we combine both lists and use sentimentr to create a dataframe with positive, negative and neutral adjectives across genders and genres.


```{r}

# Combine two lists into a single data frame
all_adjectives <- bind_rows(
  list(Mystery = mystery_adjective_list, Romance = romance_adjective_list),
  .id = "genre"
)

write.csv(all_adjectives, "all_adjectives_list.csv", row.names = FALSE)

# Create the detailed summary, grouping by both genre and gender
detailed_summary <- all_adjectives %>%
  mutate(sentiment_score = sentiment(adjective)$sentiment) %>%
  mutate(sentiment = case_when(
    sentiment_score > 0  ~ "positive",
    sentiment_score < 0  ~ "negative",
    sentiment_score == 0 ~ "neutral"
  )) %>%
  
  # The key change is here: group by both variables
  count(genre, gender, sentiment) %>%
  
  pivot_wider(names_from = sentiment, values_from = n, values_fill = list(n = 0))

# View the detailed summary table
print(detailed_summary)
```

Take unique adjectives and remove duplicates for manual revision:

```{r}
# Select unique values from the adjective column
unique_adjectives <- unique(all_adjectives$adjective)

# Convert to a data frame for CSV export
df_unique <- data.frame(adjective = unique_adjectives)

# Write to CSV
write.csv(df_unique, "unique_adjectives.csv", row.names = FALSE)
```


Contingency table for chi-squared test:

```{r}

# We unite 'genre' and 'gender' to create unique row names for the test table.
overall_contingency_table <- detailed_summary %>%
  ungroup() %>% # Ungroup to ensure unite works as expected
  unite("group", genre, gender, sep = "_") %>% # Combine genre and gender
  tibble::column_to_rownames("group") %>%      # Set the new 'group' as row names
  as.matrix()                                  # Convert to a matrix

# Let's view the final contingency table
print(overall_contingency_table)


# 2. Run the Chi-squared test on this overall table
overall_chi_test_result <- chisq.test(overall_contingency_table)

# 3. Print the results
print(overall_chi_test_result)
```
There is a significant relationship between both variables (the genre-gender groups and the sentiment categories). The way sentiments are distributed is dependent on the group. To see which specific groups and sentiments are driving this difference, we compute the standard residuals:

```{r}
overall_chi_test_result$stdres
```
And the Cramer's V to test effect sizes:

```{r}
cramerV(overall_contingency_table)
```

Cramer's V finally indicates a weak association, much like H1 results.
