---
title: "Not-So-Great-Expectations. Code for H3 testing"
author: "Rocío Galeote"
date: "2025-08-31"
output: html_document
---

## Hypothesis 3 testing

NOTE: This rmd serves as Part 5 of the accompanying code for the thesis "Not-so-Great Expectations: Analyzing Gender, Genre and Reader Bias in Commercial Fiction Using Natural Language Processing".

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval= FALSE)
```


```{r}

library(sentimentr)
library(dplyr)
library(MASS)        # polr (ordinal logistic regression)
library(lme4)        # lmer (linear mixed models)
library(ordinal)     # clmm (ordinal mixed models)
library(rsample)
library(stargazer)
library(webshot2)
library(htmltools)
library(ggplot2)
library(texreg)
library(pscl)
library(lmtest)       # Breusch–Pagan
library(car)          # VIF
library(performance)  # check_model, R2 for mixed models
library(MuMIn)        # R2 for lmer/clmm
library(brant)        # Brant test for proportional odds
library(caret)        # cross-validation
library(effects)      # marginal effects
library(ggeffects)    # marginal effect plots
```


## Mystery models

Two complementary model families were estimated: 

-sentiment score as the dependent variable, with OLS regressions: “for the same star rating, do reviewers use different language toward authors of different genders?” 

-ordinal logistic models with the inverted question as a robustness check, to assess whether the mapping from sentiment to star probability differs by gender. 

-mixed-effect regression models (1 | title) + (1 | author) to account for the fact that ultiple reviews in the datasets can be written about the same book, and many authors have several books included.

All models use the same review text sentiment scores (added with sentimentr).


```{r}

set.seed(124)

process_models_mystery <- function(data) {
  
  pipeline_start <- Sys.time()
  message("🚀 Starting Mystery models pipeline at ", pipeline_start)
  
  # --- Step 1: Sentiment scores ---
  step1_start <- Sys.time()
  message("Step 1: Computing sentiment scores...")
  
  sample_sentiment <- sentiment_by(data$review_text, by = NULL)
  
  df <- data %>%
    mutate(
      sentiment_score = sample_sentiment$ave_sentiment,
      gender = as.factor(gender),
      review_rating = factor(review_rating, ordered = TRUE)
    ) %>%
    filter(!is.na(sentiment_score), !is.na(gender), !is.na(title), !is.na(author))
  
  df$gender <- droplevels(df$gender)
  
  step1_end <- Sys.time()
  message("Step 1 complete. Duration: ", round(difftime(step1_end, step1_start, units = "secs"), 2), " secs")
  
  # --- Step 2: Linear models (sentiment outcome) ---
  step2_start <- Sys.time()
  message("Step 2: Running linear regression models (sentiment outcome)...")
  
  model_sent_base   <- lm(sentiment_score ~ review_rating, data = df)
  model_sent_gender <- lm(sentiment_score ~ review_rating + gender, data = df)
  model_sent_inter  <- lm(sentiment_score ~ review_rating * gender, data = df)
  
  step2_end <- Sys.time()
  message("Step 2 complete. Duration: ", round(difftime(step2_end, step2_start, units = "secs"), 2), " secs")
  
  # --- Step 3: Ordinal logistic models (rating outcome) ---
  step3_start <- Sys.time()
  message("Step 3: Running ordinal logistic regression models (rating outcome)...")
  
  model_rating_base   <- polr(review_rating ~ sentiment_score, data = df, Hess = TRUE)
  model_rating_gender <- polr(review_rating ~ sentiment_score + gender, data = df, Hess = TRUE)
  model_rating_inter  <- polr(review_rating ~ sentiment_score * gender, data = df, Hess = TRUE)
  
  step3_end <- Sys.time()
  message("Step 3 complete. Duration: ", round(difftime(step3_end, step3_start, units = "secs"), 2), " secs")
  
  # --- Step 4: Mixed-effects models ---
  step4_start <- Sys.time()
  message("Step 4: Running mixed-effects models (author + title random effects)...")
  
  model_sent_mixed <- lmer(sentiment_score ~ review_rating * gender + (1 | author) + (1 | title),
                           data = df,
                           control = lmerControl(optimizer = "bobyqa"))
  
  model_rating_mixed <- clmm(review_rating ~ sentiment_score * gender + (1 | author) + (1 | title),
                             data = df)
  
  step4_end <- Sys.time()
  message("Step 4 complete. Duration: ", round(difftime(step4_end, step4_start, units = "secs"), 2), " secs")
  
  pipeline_end <- Sys.time()
  message("🎉 Mystery models pipeline completed at ", pipeline_end)
  message("⏱️ Total runtime: ", round(difftime(pipeline_end, pipeline_start, units = "secs"), 2), " secs")
  
  return(list(
    # OLS sentiment models
    sent_base   = model_sent_base,
    sent_gender = model_sent_gender,
    sent_inter  = model_sent_inter,
    
    # Ordinal models
    rating_base   = model_rating_base,
    rating_gender = model_rating_gender,
    rating_inter  = model_rating_inter,
    
    # Mixed-effects models
    sent_mixed   = model_sent_mixed,
    rating_mixed = model_rating_mixed
  ))
}

# --- Run on Mystery dataset ---
models_mystery <- process_models_mystery(mystery_author_mentions)

```


Visualize results:

```{r}

# --- Linear models (sentiment outcome) ---
stargazer(
  models_mystery$sent_base, models_mystery$sent_gender, models_mystery$sent_inter,
  type = "html",
  title = "Linear Models of Sentiment Score (Outcome) ~ Rating × Gender",
  dep.var.labels = "Sentiment Score",
  column.labels = c("Base", "Gender Added", "Interaction"),
  covariate.labels = c("Review Rating", "Gender (Female)", "Gender (Unknown)", 
                       "Review Rating × Gender (Female)", "Review Rating × Gender (Unknown)"),
  omit.stat = c("f", "ser"),
  no.space = TRUE,
  out = "mystery_linear.html"
)

# Save as PNG
webshot("mystery_linear.html", file = "mystery_linear.png", vwidth = 1200, vheight = 900)


htmlreg(
  list(models_mystery$rating_base, models_mystery$rating_gender, models_mystery$rating_inter),
  file = "mystery_ordinal.html",
  custom.model.names = c("Base", "Gender Added", "Interaction"),
  caption = "Ordinal Logistic Models of Review Rating (Outcome) ~ Sentiment × Gender"
)

webshot("mystery_ordinal.html", "mystery_ordinal.png", vwidth = 1200, vheight = 900)

```



### Model validation and diagnostics

For model validation and diagnostics, OLS models were checked with residual and QQ-plots, a Breusch–Pagan test for heteroskedasticity, VIFs for multicollinearity, and 10-fold cross-validation to assess predictive performance. 

Ordinal logistic were evaluated with nominal and scale tests to test the proportional odds assumption, pseudo R², and classification accuracy. 

Lastly, mixed-effects models were assessed via marginal and conditional R², intraclass correlation coefficients, and log-likelihood. 

Marginal effects plots were additionally used to visualize predicted sentiment scores and rating probabilities across review rating and gender.

```{r}
# --- Validation & diagnostics pipeline for models_mystery ---

# Extract models
ols_model   <- models_mystery$sent_inter   # OLS interaction model
polr_model  <- models_mystery$rating_inter # Ordinal logistic interaction
lmer_model  <- models_mystery$sent_mixed   # Mixed linear
clmm_model  <- models_mystery$rating_mixed # Mixed ordinal

# --- 1. OLS diagnostics (sentiment outcome) ---
message("OLS diagnostics...")

# Residual vs Fitted
plot(ols_model, which = 1)  # should look like random scatter
# QQ-plot
plot(ols_model, which = 2)

# Breusch–Pagan test for heteroskedasticity
print(bptest(ols_model))

# VIF for multicollinearity
print(vif(ols_model))

# Cross-validation (10-fold CV, RMSE & R²)
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(
  sentiment_score ~ review_rating * gender,
  data = ols_model$model,
  method = "lm",
  trControl = train_control
)
print(cv_model)

# --- 2. Ordinal logistic diagnostics (rating outcome) ---
message("Ordinal logistic regression diagnostics...")

# Brant test (proportional odds assumption)
print(brant(polr_model))

# Pseudo R² measures
print(pR2(polr_model))

# Classification accuracy on observed vs predicted
pred_class <- predict(polr_model, type = "class")
acc <- mean(pred_class == polr_model$model$review_rating)
message("Classification accuracy: ", round(acc, 3))

# --- 3. Mixed-effects models (sample-based) ---
message("Mixed-effects models diagnostics...")

# lmer: R²
print(r.squaredGLMM(lmer_model))

# Check ICC to justify random effects
icc_val <- performance::icc(lmer_model)
print(icc_val)

# clmm: no direct R², but logLik can be reported
logLik(clmm_model)

# --- 4. Marginal effects / predicted probabilities ---
message("Computing marginal effects and plotting interactions...")

# Marginal effects for OLS (sentiment outcome)
eff_ols <- ggpredict(ols_model, terms = c("review_rating", "gender"))
plot(eff_ols) + ggtitle("Predicted Sentiment by Rating × Gender")

# Marginal effects for ordinal logistic (rating outcome)
eff_polr <- ggpredict(polr_model, terms = c("sentiment_score", "gender"))
plot(eff_polr) + ggtitle("Predicted Rating Probabilities by Sentiment × Gender")

# --- 5. Save outputs (optional) ---
ggsave("mystery_sentiment_interaction.png", plot = last_plot(), width = 8, height = 5, dpi = 300)

```

The OLS model for sentiment shows low multicollinearity (adjusted GVIFs < 2) but significant heteroskedasticity (BP = 228.84, p < 0.001). Predictive performance is modest, with R² ≈ 0.089, RMSE ≈ 0.171, and MAE ≈ 0.128. The model explains about 9% of the variance in sentiment scores. It detects patterns between sentiment, star rating, and gender, but effect sizes are small and heteroskedasticity should be considered.

The mixed-effects models indicate that fixed effects explain ~9% of variance in sentiment (marginal R² = 0.094), while including random effects for book title increases explained variance to ~15% (conditional R² = 0.152). ICC = 0.065 shows that roughly 6–7% of the variance is attributable to clustering by title. 

For the ordinal mixed model, log-likelihood confirms model fit. Overall, accounting for hierarchical structure improves explanatory power, though most variance remains at the individual review level.


## Romance models

Compute sentiment score, weigh the dataset in order to balance genders, and then compute linear, ordinal logistic and mixed-effects models:

```{r}

set.seed(124)

process_models <- function(data) {
  
  # --- 1. Compute sentiment scores ---
  sample_sentiment <- sentiment_by(data$review_text, by = NULL)
  
  df <- data %>%
    mutate(
      sentiment_score = sample_sentiment$ave_sentiment,
      gender = as.factor(gender),
      review_rating = factor(review_rating, ordered = TRUE)
    ) %>%
    filter(!is.na(sentiment_score), !is.na(gender), !is.na(title))
  
  df$gender <- droplevels(df$gender)
  
  # --- 2. Weighting scheme (to balance genders) ---
  weights <- ifelse(df$gender == "male", 
                    nrow(df) / (2 * sum(df$gender == "male")), 
                    ifelse(df$gender == "female",
                           nrow(df) / (2 * sum(df$gender == "female")), 1))
  
  # --- 3. Weighted models (MAIN analysis) ---
  # Sentiment outcome
  model_sent_base_w   <- lm(sentiment_score ~ review_rating, data = df, weights = weights)
  model_sent_gender_w <- lm(sentiment_score ~ review_rating + gender, data = df, weights = weights)
  model_sent_inter_w  <- lm(sentiment_score ~ review_rating * gender, data = df, weights = weights)
  
  # Ordinal logistic outcome
  model_rating_base_w   <- polr(review_rating ~ sentiment_score, data = df, weights = weights, Hess = TRUE)
  model_rating_gender_w <- polr(review_rating ~ sentiment_score + gender, data = df, weights = weights, Hess = TRUE)
  model_rating_inter_w  <- polr(review_rating ~ sentiment_score * gender, data = df, weights = weights, Hess = TRUE)
  
  # --- 4. Downsampled dataset (ROBUSTNESS check) ---
  male_df    <- df %>% filter(gender == "male")
  female_df  <- df %>% filter(gender == "female")
  unknown_df <- df %>% filter(gender == "unknown")
  
  set.seed(124)
  female_down <- female_df %>% sample_n(min(nrow(male_df), nrow(female_df)))
  
  balanced_df <- bind_rows(male_df, female_down, unknown_df)
  balanced_df$gender <- droplevels(balanced_df$gender)
  
  # Sentiment outcome
  model_sent_base_bal   <- lm(sentiment_score ~ review_rating, data = balanced_df)
  model_sent_gender_bal <- lm(sentiment_score ~ review_rating + gender, data = balanced_df)
  model_sent_inter_bal  <- lm(sentiment_score ~ review_rating * gender, data = balanced_df)
  
  # Ordinal logistic outcome
  model_rating_base_bal   <- polr(review_rating ~ sentiment_score, data = balanced_df, Hess = TRUE)
  model_rating_gender_bal <- polr(review_rating ~ sentiment_score + gender, data = balanced_df, Hess = TRUE)
  model_rating_inter_bal  <- polr(review_rating ~ sentiment_score * gender, data = balanced_df, Hess = TRUE)
  
  # --- 5. Mixed-effects models (account for clustering by title) ---
  # Linear mixed model (sentiment outcome)
  model_sent_mixed <- lmer(sentiment_score ~ review_rating * gender + (1 | title),
                           data = df, weights = weights,
                           control = lmerControl(optimizer = "bobyqa"))
  
  # Ordinal mixed model (rating outcome)
  model_rating_mixed <- clmm(review_rating ~ sentiment_score * gender + (1 | title),
                             data = df, weights = weights)
  
  return(list(
    # Weighted models
    sent_base_w   = model_sent_base_w,
    sent_gender_w = model_sent_gender_w,
    sent_inter_w  = model_sent_inter_w,
    rating_base_w   = model_rating_base_w,
    rating_gender_w = model_rating_gender_w,
    rating_inter_w  = model_rating_inter_w,
    
    # Downsampled models
    sent_base_bal   = model_sent_base_bal,
    sent_gender_bal = model_sent_gender_bal,
    sent_inter_bal  = model_sent_inter_bal,
    rating_base_bal   = model_rating_base_bal,
    rating_gender_bal = model_rating_gender_bal,
    rating_inter_bal  = model_rating_inter_bal,
    
    # Mixed-effects models
    sent_mixed   = model_sent_mixed,
    rating_mixed = model_rating_mixed
  ))
}

# --- Run on Romance dataset (replace with Mystery dataset as needed) ---
models_romance <- process_models(romance_author_mentions)

# --- Example Stargazer output: Linear models (sentiment outcome) ---
stargazer(
  models_romance$sent_base_w, models_romance$sent_gender_w, models_romance$sent_inter_w,
  models_romance$sent_base_bal, models_romance$sent_gender_bal, models_romance$sent_inter_bal,
  type = "text",
  title = "Linear Models of Sentiment Score (Outcome) ~ Rating × Gender"
)

# Mixed models need to be summarized separately
summary(models_romance$sent_mixed)
summary(models_romance$rating_mixed)

```

Now we visualize them with Stargazer and htmlreg:

```{r}

# --- Linear models (sentiment outcome) ---
linear_models <- list(
  models_romance$sent_base_w,
  models_romance$sent_gender_w,
  models_romance$sent_inter_w,
  models_romance$sent_base_bal,
  models_romance$sent_gender_bal,
  models_romance$sent_inter_bal
)

# Drop NULLs or bad models just in case
linear_models <- Filter(Negate(is.null), linear_models)

stargazer(
  linear_models,
  type = "html",
  title = "Linear Models of Sentiment Score (Outcome) ~ Rating × Gender",
  dep.var.labels = "Sentiment Score",
  column.labels = c("Base (W)", "Gender Added (W)", "Interaction (W)",
                    "Base (Balanced)", "Gender Added (Balanced)", "Interaction (Balanced)"),
  covariate.labels = c("Review Rating", "Gender (Female)", "Gender (Unknown)", 
                       "Review Rating × Gender (Female)", "Review Rating × Gender (Unknown)"),
  omit.stat = c("f", "ser"),
  no.space = TRUE,
  out = "romance_linear.html"
)

# Save as PNG
webshot("romance_linear.html", file = "romance_linear.png", vwidth = 1200, vheight = 900)


# --- Ordinal logistic models (review rating outcome) ---
rating_models <- list(
  extract(models_romance$rating_base_w),
  extract(models_romance$rating_gender_w),
  extract(models_romance$rating_inter_w),
  extract(models_romance$rating_base_bal),
  extract(models_romance$rating_gender_bal),
  extract(models_romance$rating_inter_bal)
)

htmlreg(
  rating_models,
  file = "romance_ordinal.html",
  custom.model.names = c("Base (W)", "Gender Added (W)", "Interaction (W)",
                         "Base (Balanced)", "Gender Added (Balanced)", "Interaction (Balanced)"),
  caption = "Ordinal Logistic Models of Review Rating (Outcome) ~ Sentiment × Gender"
)

webshot("romance_ordinal.html", "romance_ordinal.png", vwidth = 1200, vheight = 900)


# --- Mixed models (summarize separately) ---
summary(models_romance$sent_mixed)
summary(models_romance$rating_mixed)


```

Code for mixed effects:

```{r}

# Extract tidy summaries
sent_mixed_tbl <- tidy(models_romance$sent_mixed, effects = "fixed") %>%
  mutate(model = "Sentiment Mixed")

rating_mixed_tbl <- tidy(models_romance$rating_mixed, effects = "fixed") %>%
  mutate(model = "Rating Mixed")

# Combine into one dataframe
mixed_tbl <- bind_rows(sent_mixed_tbl, rating_mixed_tbl)

# Show as table in console
print(mixed_tbl)

# Export to HTML
kable(mixed_tbl, format = "html", caption = "Fixed Effects from Mixed Models") %>%
  cat(file = "romance_mixed.html")

webshot("romance_mixed.html", "romance_mixed.png", vwidth = 1200, vheight = 900)

```


### Model validation and diagnostics

For model validation and diagnostics, OLS models were checked with residual and QQ-plots, a Breusch–Pagan test for heteroskedasticity, VIFs for multicollinearity, and 10-fold cross-validation to assess predictive performance. 

Ordinal logistic were evaluated with nominal and scale tests to test the proportional odds assumption, pseudo R², and classification accuracy. 

Lastly, mixed-effects models were assessed via marginal and conditional R², intraclass correlation coefficients, and log-likelihood. 

Marginal effects plots were additionally used to visualize predicted sentiment scores and rating probabilities across review rating and gender.

```{r}

# --- Validation & diagnostics pipeline for models_romance ---

# Extract models
ols_model   <- models_romance$sent_inter_w   # Weighted OLS interaction model
polr_model  <- models_romance$rating_inter_w # Weighted ordinal logistic interaction
lmer_model  <- models_romance$sent_mixed     # Mixed linear
clmm_model  <- models_romance$rating_mixed   # Mixed ordinal

# --- 1. OLS diagnostics (sentiment outcome) ---
message("OLS diagnostics...")

# Residual vs Fitted
plot(ols_model, which = 1)  # should look like random scatter
# QQ-plot
plot(ols_model, which = 2)

# Breusch–Pagan test for heteroskedasticity
print(bptest(ols_model))

# VIF for multicollinearity
print(vif(ols_model))

# Cross-validation (10-fold CV, RMSE & R²)
set.seed(123)
train_control <- trainControl(method = "cv", number = 10)
cv_model <- train(
  sentiment_score ~ review_rating * gender,
  data = ols_model$model,
  method = "lm",
  trControl = train_control
)
print(cv_model)

# --- 2. Ordinal logistic diagnostics (rating outcome) ---
message("Ordinal logistic regression diagnostics...")


# Use clm from 'ordinal' instead of polr
clm_model <- ordinal::clm(
  review_rating ~ sentiment_score * gender,
  data = models_romance$rating_inter_w$model,
  weights = models_romance$rating_inter_w$weights,
  link = "logit"
)

# Nominal test (like Brant test)
nominal_test(clm_model)

# Scale test (detects non-proportionality)
scale_test(clm_model)
# Pseudo R² measures
print(pR2(polr_model))

# Classification accuracy on observed vs predicted
# Get predicted classes
pred_class <- predict(polr_model, type = "class")

# Convert both predicted and true to character
true_class <- as.character(polr_model$model$review_rating)
pred_class <- as.character(pred_class)

# Compute accuracy
acc <- mean(pred_class == true_class)
message("Classification accuracy: ", round(acc, 3))


# --- 3. Mixed-effects models (sample-based) ---
message("Mixed-effects models diagnostics...")

# lmer: R²
print(r.squaredGLMM(lmer_model))

# Check ICC to justify random effects
icc_val <- performance::icc(lmer_model)
print(icc_val)

# clmm: no direct R², but logLik can be reported
print(logLik(clmm_model))

# --- 4. Marginal effects / predicted probabilities ---
message("Computing marginal effects and plotting interactions...")


# Marginal effects for OLS (sentiment outcome)
eff_ols <- ggpredict(ols_model, terms = c("review_rating", "gender"))
plot(eff_ols) + ggtitle("Predicted Sentiment by Rating × Gender")

# Marginal effects for ordinal logistic (rating outcome)
eff_polr <- ggpredict(polr_model, terms = c("sentiment_score", "gender"))
plot(eff_polr) + ggtitle("Predicted Rating Probabilities by Sentiment × Gender")

# --- 5. Save outputs (optional) ---
ggsave("romance_sentiment_interaction.png", plot = last_plot(), width = 8, height = 5, dpi = 300)

```

Results indicate that sentiment scores and review ratings are influenced by both sentiment, gender, and their interaction, but with varying degrees of explanatory power across model types. 

The OLS interaction model shows that predictors are not highly correlated (VIFs < 2), but the Breusch–Pagan test detects heteroskedasticity, and cross-validated performance is modest (R² ≈ 0.11, RMSE ≈ 0.15), meaning limited predictive power. 

Ordinal logistic model reveals significant effects of sentiment, gender, and their interaction, but nominal and scale tests indicate that the proportional odds assumption is violated.

Mixed-effects models show that accounting for title-level clustering improves fit. 

Overall, while all predictors have significant relationships with outcomes, effect sizes and predictive performance are modest. Results should be interpreted with caution  due to heteroskedasticity, non-proportional odds, and residual variance.